Notebook: Predictors of frog survival following translocation

### 2023-03-29 doc/notebook/notebook_translocation.Rmd

* site_id: site to which frogs were translocated  
* elevation: water body surface elevation above sea level, in meters  
* shore: presence/absence of shore/bank habitat available to frogs in winter (estimated in late-summer)  
* snow_t: snow water content (measured on April 1) in year of translocation  
* snow_t1: snow water content (measured on April 1) in year following translocation  
* date: date frogs released at recipient site  
* year: year frogs released at recipient site  
* day: day of the year on which frogs were released  
* order: order of translocations to a site, from 1-4. Re-coded as indicator variable (variable name = "first"): 1 = first translocation to site; 0 = subsequent translocation  
* donor: site_id from which frogs were collected for translocation  
* pit_tag_ref: pit tag id  
* sex: m or f, re-coded as indicator variable (variable name = "male"): 1 = male, 0 = female   
* length: frog snout-vent length at capture at donor site (in mm)  
* weight: frog weight at capture at donor site (in grams)  
* bd_load: number of ITS copies at release, measured from swab collected at capture at donor site  
* survival: individual-level estimated probability of survival 1 year following translocation, based on cmr surveys conducted for at least two years post-translocation. Re-coded as indicator variable (variable name = "surv"): 0 = survival probability < 0.5; 1 = survival probability $\geq$ 0.5)

Other variables
* Maximum depth and surface area would seem useful to include, but 74976 is a stream/meadow habitat and as such depth and area at that site are likely not comparable to depth and area at lake habitats.  
* Frog condition is problematic because size and residuals from length:weight regression line are correlated. Specifically, the largest frogs have the largest positive residuals from a length:weight regression line, even after log10 transformation. As such, length and condition are not independent.

@@

* Another consideration when designing the analysis is that there are 12 records in which bd_load is null. Of these, 9 are from 70413 on 2013-07-15, likely due to a rush to get frogs packaged prior to helicopter arrival. 39 frogs were translocated on that date, so plenty of swab results are available to represent that site. Prior to running the analyses, should drop the 12 records where bd_load = NA. If bd_load is not an important predictor, can run analysis (without the bd_load predictor) using the full dataset.

@@

Exploratory data analyses conducted in translocation_survival_exploratory.Rmd provided several notable insights.

### Frog survival across recipient sites
* 1-year frog survival is highly variable across sites, but highly repeatable within a site. This indicates the importance of including site_id or translocation_id as group-level effects. It remains unclear whether including a translocation_id nested within site_id should be considered (or if this nested group-level effect is even feasible given the relatively small number of translocations per site).

### Donor site
* Donor site has a strong association with frog survival, with survival at 70459 > 70567 > 72996. Explored possible effects of bd_load and frog size as drivers of this pattern. Differences in survival between donor sites seem unrelated to bd_load, but appear closely associated with frog size. Whether frog size is the sole driver or is a result of frog size plus other unmeasured donor site attributes (e.g., degree of resistance against Bd infection) is unknown.  
  * Minimal differences between donor sites in median bd_load, but slightly higher at 72996. No obvious effect of bd_load on 1-year frog survival.  
  * Strong effect of frog size on 1-year survival, with frogs > 55 mm having substantially higher 1-year survival. Frogs >60 mm had by far the highest survival.  
  * The effect of frog size on survival appears unrelated to bd_load because there is no relationship between frog size and bd_load.  
  * Frog size differs between donor site: size at 72996 < 70459 = 70567. This partially mirrors the observed differences in survival between sites. Sufficient overlap in frog sizes between donor sites that it might be possible to include both frog length and donor in the model.

### Bd load
* Frogs with the highest loads at the time of translocation had the lowest survival. The effect was only evident in the category that included frogs with loads of 56,000--2,100,000 ITS copies. Notable result given that the threshold level of 56,000 copies is /~ 1 order of magnitude below the 600,000 copies level associated with frog mortality during epizootics.

### Recipient site characteristics
* **Shore habitat**: The presence of shore habitat is associated with increased frog survival. Unknown if this is a direct effect reflecting the importance of this habitat type, or if shore habitat is serving as a proxy for another habitat feature. Tight association between shore habitat presence/absence and elevation will make it impossible to disentangle these effects, include only elevation in the model.  
* **Elevation**: Strong positive effect on frog survival.

### Translocation timing
* Day of the year showed no obvious relationship to frog survival.  
* Year was positively associated with survival. This is likely not a direct effect of year, but instead reflects better-informed selection of recipient sites with high-quality frog habitat over time. As such, consider not including year in the model.

### Winter severity
* No obvious relationship between either winter severity in year of translocation or in year following translocation on frog survival, but some substantial differences in frog survival between winter severity categories.

### Translocation order
* Frogs in the first cohort translocated to a site generally had higher survival than frogs in subsequent cohorts. However, effect appears relatively weak.  
* It remains unclear what mechanism could lead to the slightly higher mortality of later cohorts.

### Frog characteristics
* Frog size had strong effect on 1-year survival.  
* Minimal differences in survival between males and females (slightly higher in males).

@@

Pairs plot to identify correlations between predictors (ggpairs from GGally package):
* It may not be possible to include both donor and length in a single model because the distribution of frog lengths is quite different across donor sites, causing dependency between donor and length. As such, these two variables have the potential to compete for the same variation. Need to look at this more closely, perhaps by plotting frog lengths at each donor site. Also, plot length against survival, grouped by donor.  Are these distributions as donor-specific as they appear to be? Because of the importance of distinguishing frog size effects from other effects associated with donor populations (e.g., inherent differences in Bd susceptibility), may need to explore whether it would be possible to compare the fit of two models, each containing one of these two variables.  
* Also need to disentangle elevation from site_id - currently only one elevation per site_id. Try breaking continuous elevation into categories, with > 2 site_ids per category. Hopefully, that will make elevation variable less dependent on site_id. A similar situation may exist for the two winter severity variables and site_id.  
* Shore and elevation may also be closely associated. Are all sites with shore = 0 at low elevation? If so, may need to only include elevation in model.

@@

* Bin elevation into "low" and "high" groups.  
* Recode order as 0/1 (instead of 1--4): "first".  
* Recode survival as 0/1 (instead of continuous probability): "surv".  
* Recode binary variables as 0/1 indicators: elevation ("elev_lo"), sex ("male"). See Gelman reference below.  
* Format relevant variables as factors: surv, elev_lo, shore, first, donor, male, year, site_id, translocation_id.  
* Standardize continuous variables: elevation, snow_t, snow_t1, day, length. Possibly standardize log(bd_load) also, but see Gelman for recommendations. Gelman recommends centering by subtracting the mean and standardizing by dividing by 2 standard deviations. Centering to a mean of 0 makes interactions more interpretable. Standardizing allows assessment of importance of predictors by comparing regression coefficients. Dividing by 2 SD puts continuous variables on approximately the same scale as binary (0/1) variables. This is true only if binary predictors have similar number of 0s and 1s. In the current dataset, there is some deviation from 1:1. Sex: female = 428, male = 351; order: 0 = 507, 1 = 272. Binary variables should be centered but not standardized. Variables in need of log-transformation should be transformed and then rescaled, although rescaling will not always be necessary.

Relevant references:  
* Gelman 2008: https://doi.org/10.1002/sim.3107  
* Schielzeth 2010: https://doi.org/10.1111/j.2041-210X.2010.00012.x

### Priors
* For regression coefficients in logistic regression: A recommended weakly informative prior is beta ~ student_t(nu,0,s), where s is chosen to provide weak information on the expected scale, and 3<nu<7.  https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations.  
* Other recommendations in:  
  * Moving beyond noninformative priors: why and how to choose weakly informative priors in Bayesian analyses: https://doi.org/10.1111/oik.05985.  
  * Choosing priors in Bayesian ecological models by simulating from the prior predictive distribution: https://doi.org/10.1002/ecs2.3739.

## Results
- Important predictors of the 1-year survival of translocated frogs were quite consistent across all models (see `translocation_analysis_survival.Rmd` for details).  
- In the "best" model that included standardized predictors and translocation_id as a grouping variable, important predictors included elevation (+), winter severity in the year following translocation (-), and donor[72996] (-). Model R2 was 0.32, so a substantial amount of variation remained unaccounted for.  
- Comparison of different grouping structures (translocation_id, site_id, translocation_id nested within site_id) indicated that all had similar posterior predictive accuracy. However, using translocation_id as the grouping variable allowed greater insights into the effects of the predictor variables, presumably due to the greater variation associated with the translocation_id grouping variable (n = 24) compared to the site_id grouping variables (n = 12).  
- Posterior predictive accuracy of "best" model was 0.79, and the true negative rate (proportion of frogs that did not survive that were accurately classified as not surviving: 0.88) was higher than the true positive rate (proportion of frogs that survived that were accurately classified as surviving: 0.59). Indicates that with the available site-level and frog-level predictors, frog non-survival is more accurately predicted than frog survival. This is borne out by sites such as 70370 which has attributes suggesting high frog survival (e.g., high elevation, available bank habitat, few predators), but in fact frogs from both translocations experienced very low survival. In contrast, no examples exist of sites where frog survival was (retrospectively) predicted to be low, but actual survival was high (although 70449 comes closest to providing such an example).   
- Two supplementary analyses were conducted to gain additional insights into the 1-year survival model.  
  - Because the three donor populations differed in the size of collected/translocated frogs, frog size was potentially associated with the important effect of donor population. Therefore, I reran the analysis using only frogs translocated from 72996. Despite substantial variation in the size of frogs collected/translocated from this site, results indicated that frog size was unrelated to frog survival. This suggests that the effect of donor population is due to an unmeasured population-level attribute, e.g., a characteristic associated with frog resistance/tolerance of Bd infection. The existence of such an attribute is suggested by the consistently high survival resulting from the four translocations conducted to date from 70459 to two very different sites, and the fact that in both recipient sites that have received frogs from 70459 tadpole mouthparts are fully pigmented. That is not the case with tadpoles at sites that received frogs from 72996.    
  - The frog translocation dataset contained 12 records for which bd_load = NA. To allow inclusion of bd_load in the model, these 12 records were dropped. Given that bd_load was not an important predictor of frog survival, I reran the model using the complete dataset and dropping bd_load from the model. Dropping/including those 12 records had no effect on model results.   
- The results from a secondary analysis to complement the 1-yr survival model demonstrated that the average survival in the first translocation was an important predictor of frog survival in subsequent translocations. This demonstrates the importance of recipient site attributes (elevation, habitat, predators, etc.) in determining frog survival.

>>>>

-   site_id: site to which frogs were translocated

-   elevation: water body surface elevation above sea level, in meters

-   shore: presence/absence of shore/bank habitat available to frogs in winter (estimated in late-summer)

-   snow_t: snow water content (measured on April 1) in year of translocation

-   snow_t1: snow water content (measured on April 1) in year following translocation

-   date: date frogs released at recipient site

-   year: year frogs released at recipient site

-   day: day of the year on which frogs were released

-   order: order of translocations to a site, from 1-4.
    Re-coded as indicator variable (variable name = "first"): 1 = first translocation to site; 0 = subsequent translocation

-   donor: site_id from which frogs were collected for translocation

-   pit_tag_ref: pit tag id

-   sex: m or f, re-coded as indicator variable (variable name = "male"): 1 = male, 0 = female

-   length: frog snout-vent length at capture at donor site (in mm)

-   weight: frog weight at capture at donor site (in grams)

-   bd_load: number of ITS copies at release, measured from swab collected at capture at donor site

-   survival: individual-level estimated probability of survival 1 year following translocation, based on cmr surveys conducted for at least two years post-translocation.
    Re-coded as indicator variable (variable name = "surv"): 0 = survival probability \< 0.5; 1 = survival probability $\geq$ 0.5)

Other variables:

-   Maximum depth and surface area would seem useful to include, but 74976 is a stream/meadow habitat and as such depth and area at that site are likely not comparable to depth and area at lake habitats.

-    Frog condition is problematic because size and residuals from length:weight regression line are correlated.
    Specifically, the largest frogs have the largest positive residuals from a length:weight regression line, even after log10 transformation.
    As such, length and condition are not independent.

@@

-   Another consideration when designing the analysis is that there are 12 records in which bd_load is null. Of these, 9 are from 70413 on 2013-07-15, likely due to a rush to get frogs packaged prior to helicopter arrival. 39 frogs were translocated on that date, so plenty of swab results are available to represent that site. Prior to running the analyses, should drop the 12 records where bd_load = NA. If bd_load is not an important predictor, can run analysis (without the bd_load predictor) using the full dataset.

@@

Exploratory data analyses of univariate associations between predictors and frog survival provided several insights.

### Frog survival across recipient sites

-   1-year frog survival is highly variable across sites, but highly repeatable within a site. This indicates the importance of including site_id or translocation_id as group-level effects. It remains unclear whether including a translocation_id nested within site_id should be considered (or if this nested group-level effect is even feasible given the relatively small number of translocations per site).

### Donor site

-   Donor site has a strong association with frog survival, with survival at 70459 \> 70567 \> 72996. Explored possible effects of bd_load and frog size as drivers of this pattern. Differences in survival between donor sites seem unrelated to bd_load, but appear closely associated with frog size. Whether frog size is the sole driver or is a result of frog size plus other unmeasured donor site attributes (e.g., degree of resistance against Bd infection) is unknown.
    -   Minimal differences between donor sites in median bd_load, but slightly higher at 72996. No obvious effect of bd_load on 1-year frog survival.
    -   Strong effect of frog size on 1-year survival, with frogs \> 55 mm having substantially higher 1-year survival. Frogs \>60 mm had by far the highest survival.
    -   The effect of frog size on survival appears unrelated to bd_load because there is no relationship between frog size and bd_load.
    -   Frog size differs between donor site: size at 72996 \< 70459 = 70567. This partially mirrors the observed differences in survival between sites. Sufficient overlap in frog sizes between donor sites that it might be possible to include both frog length and donor in the model.

### Bd load

-   Frogs with the highest loads at the time of translocation had the lowest survival. The effect was only evident in the category that included frogs with loads of 56,000--2,100,000 ITS copies. Notable result given that the threshold level of 56,000 copies is /\~ 1 order of magnitude below the 600,000 copies level associated with frog mortality during epizootics.

### Recipient site characteristics

-   Shore habitat: The presence of shore habitat is associated with increased frog survival. Unknown if this is a direct effect reflecting the importance of this habitat type, or if shore habitat is serving as a proxy for another habitat feature. Tight association between shore habitat presence/absence and elevation will make it impossible to disentangle these effects, include only elevation in the model.
-   Elevation: Strong positive effect on frog survival.

### Translocation timing

-   Day of the year showed no obvious relationship to frog survival.\
-   Year was positively associated with survival. This is likely not a direct effect of year, but instead reflects better-informed selection of recipient sites with high-quality frog habitat over time. As such, consider not including year in the model.

### Winter severity

-   No obvious relationship between either winter severity in year of translocation or in year following translocation on frog survival, but some substantial differences in frog survival between winter severity categories.

### Translocation order

-   Frogs in the first cohort translocated to a site generally had higher survival than frogs in subsequent cohorts. However, effect appears relatively weak.
-   It remains unclear what mechanism could lead to the slightly higher mortality of later cohorts.

### Frog characteristics

-   Frog size is associated with 1-year survival.
-   Minimal differences in survival between males and females (slightly higher in males).

@@

Pairs plot to identify correlations between predictors (ggpairs from GGally package):

-   It may not be possible to include both donor and length in a single model because the distribution of frog lengths is quite different across donor sites, causing dependency between donor and length.
    As such, these two variables have the potential to compete for the same variation.
    Need to look at this more closely, perhaps by plotting frog lengths at each donor site.
    Also, plot length against survival, grouped by donor.
    Are these distributions as donor-specific as they appear to be?
    Because of the importance of distinguishing frog size effects from other effects associated with donor populations (e.g., inherent differences in Bd susceptibility), may need to explore whether it would be possible to compare the fit of two models, each containing one of these two variables.

-   Also need to disentangle elevation from site_id - currently only one elevation per site_id.
    Try breaking continuous elevation into categories, with \> 2 site_ids per category.
    Hopefully, that will make elevation variable less dependent on site_id.
    A similar situation may exist for the two winter severity variables and site_id.

-   Shore and elevation are also closely associated.
    Are all sites with shore = 0 at low elevation?
    If so, may need to only include elevation in model.

@@

-   Bin elevation into "low" and "high" groups.\
-   Recode order as 0/1 (instead of 1--4): "first".\
-   Recode survival as 0/1 (instead of continuous probability): "surv".\
-   Recode binary variables as 0/1 indicators: elevation ("elev_lo"), sex ("male"). See Gelman reference below.\
-   Format relevant variables as factors: surv, elev_lo, shore, first, donor, male, year, site_id, translocation_id.\
-   Standardize continuous variables: elevation, snow_t, snow_t1, day, length. Possibly standardize log(bd_load) also, but see Gelman for recommendations. Gelman recommends centering by subtracting the mean and standardizing by dividing by 2 standard deviations. Centering to a mean of 0 makes interactions more interpretable. Standardizing allows assessment of importance of predictors by comparing regression coefficients. Dividing by 2 SD puts continuous variables on approximately the same scale as binary (0/1) variables. This is true only if binary predictors have similar number of 0s and 1s. In the current dataset, there is some deviation from 1:1. Sex: female = 428, male = 351; order: 0 = 507, 1 = 272. Binary variables should be centered but not standardized. Variables in need of log-transformation should be transformed and then rescaled, although rescaling will not always be necessary.

Relevant references:\
\* Gelman 2008: <https://doi.org/10.1002/sim.3107>\
\* Schielzeth 2010: <https://doi.org/10.1111/j.2041-210X.2010.00012.x>

### Priors

-   For regression coefficients in logistic regression: A recommended weakly informative prior is beta \~ student_t(nu,0,s), where s is chosen to provide weak information on the expected scale, and 3\<nu\<7. <https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations>.\
-   Other recommendations in:
    -   Moving beyond noninformative priors: why and how to choose weakly informative priors in Bayesian analyses: <https://doi.org/10.1111/oik.05985>.\
    -   Choosing priors in Bayesian ecological models by simulating from the prior predictive distribution: <https://doi.org/10.1002/ecs2.3739>.

## Results

-   Important predictors of the 1-year survival of translocated frogs were quite consistent across all models (see `translocation_analysis_survival.Rmd` for details).\
-   In the "best" model that included standardized predictors and translocation_id as a grouping variable, important predictors included elevation (+), winter severity in the year following translocation (-), and donor[72996] (-). Model R2 was 0.32, so a substantial amount of variation remained unaccounted for.\
-   Comparison of different grouping structures (translocation_id, site_id, translocation_id nested within site_id) indicated that all had similar posterior predictive accuracy. However, using translocation_id as the grouping variable allowed greater insights into the effects of the predictor variables, presumably due to the greater variation associated with the translocation_id grouping variable (n = 24) compared to the site_id grouping variables (n = 12).\
-   Posterior predictive accuracy of "best" model was 0.79, and the true negative rate (proportion of frogs that did not survive that were accurately classified as not surviving: 0.88) was higher than the true positive rate (proportion of frogs that survived that were accurately classified as surviving: 0.59). Indicates that with the available site-level and frog-level predictors, frog non-survival is more accurately predicted than frog survival. This is borne out by sites such as 70370 which has attributes suggesting high frog survival (e.g., high elevation, available bank habitat, few predators), but in fact frogs from both translocations experienced very low survival. In contrast, no examples exist of sites where frog survival was (retrospectively) predicted to be low, but actual survival was high (although 70449 comes closest to providing such an example).\
-   Two supplementary analyses were conducted to gain additional insights into the 1-year survival model.
    -   Because the three donor populations differed in the size of collected/translocated frogs, frog size was potentially associated with the important effect of donor population. Therefore, I reran the analysis using only frogs translocated from 72996. Despite substantial variation in the size of frogs collected/translocated from this site, results indicated that frog size was unrelated to frog survival. This suggests that the effect of donor population is due to an unmeasured population-level attribute, e.g., a characteristic associated with frog resistance/tolerance of Bd infection. The existence of such an attribute is suggested by the consistently high survival resulting from the four translocations conducted to date from 70459 to two very different sites, and the fact that in both recipient sites that have received frogs from 70459 tadpole mouthparts are fully pigmented. That is not the case with tadpoles at sites that received frogs from 72996.\
    -   The frog translocation dataset contained 12 records for which bd_load = NA. To allow inclusion of bd_load in the model, these 12 records were dropped. Given that bd_load was not an important predictor of frog survival, I reran the model using the complete dataset and dropping bd_load from the model. Dropping/including those 12 records had no effect on model results.\
-   The results from a secondary analysis to complement the 1-yr survival model demonstrated that the average survival in the first translocation was an important predictor of frog survival in subsequent translocations. This demonstrates the importance of recipient site attributes (elevation, habitat, predators, etc.) in determining frog survival.

### Visualize conditional effects

-   As described in <https://rdrr.io/cran/brms/man/conditional_effects.brmsfit.html>, "When creating conditional_effects for a particular predictor (or interaction of two predictors), one has to choose the values of all other predictors to condition on. By default, the mean is used for continuous variables and the reference category is used for factors, but you may change these values via argument conditions. This also has an implication for the points argument: In the created plots, only those points will be shown that correspond to the factor levels actually used in the conditioning, in order not to create the false impression of bad model fit, where it is just due to conditioning on certain factor levels."
-   Created a new dataset for each predictor in which all variables except the one of interest were set to mean values (continuous variables) and the reference level (categorical variables).
